SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2013-01-14 paper/paper.tex
paper: proofreading changes.

\begin{abstract}
QuickCheck is a powerful library for automatic test-case generation.  Because
QuickCheck performs random testing, it can discover very large counterexamples,
even when enumerating test-cases is infeasible, as is done with SmallCheck.
QuickCheck provides an interface for the user to write a \emph{shrink} function
to attempt to reduce the size of a counterexample discovered by QuickCheck.
Implementations of \emph{shrink} often require significant boilerplate and are
not ``smart'', just generating structurally smaller values.  We introduce
\emph{SmartCheck}, an automatic efficient counterexample reduction algorithm.
SmartCheck reduces algebraic data using generic search heuristics to efficiently
find smaller counterexamples.  In addition to shrinking, SmartCheck generalizes
counterexamples by returning a formula that abstracts over individual
counterexamples.  SmartCheck has been implemented for Haskell and tested on a
variety of Haskell programs.
\end{abstract}

@@

\todo{define variant, tag}
\todo{note constants are user-defined arguments, other simplifications}

@@

-- SmallCheck instances
instance Serial Int16 where
  series d = drawnFrom [(-d')..d']
    where d' = fromIntegral d

instance Serial A where series = cons1 A
instance Serial B where series = cons4 B

@@

While \ttp{shrink} is user-defined and can have any type-correct implementation,
it is typically defined using structural recursion for composite data types.
For example, consider the \ttp{shrink} instances given to data types \ttp{A} and
\ttp{B} in Figure~\ref{fig:initial}.\footnote{All examples and algorithms in
  this paper are presented in Haskell~\cite{haskell98}.}  Indeed, because of the regular
structure of \ttp{shrink} implementations, the principal motivation for one of
Haskell's first generics package was for providing generic definitions for
\ttp{shrink}~\cite{syb}.

However, this straightforward approach to defining \ttp{shrink} instances may
either miss minimal counterexamples, be too inefficient to be practical, or
both.\footnote{For a recent real-world case, see the user question posed on
  StackOverflow:
  \url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}.}
For example, consider again Figure~\ref{fig:initial}.  The data type \ttp{B} is
a product type over lists of 16-bit signed integers, wrapped by a constructor.
Consider the property \ttp{prop} in the figure, stating that if the sum of each
list in a field of \ttp{B} is less than 16, then the sum of the four fields is
less than 64.  The property \ttp{prop} seems reasonable at first glance, until
one realizes that due to underflow, the property can be violated.  For example,
consider the value
%
\begin{code}
B [A (-32769)] [] [] []
\end{code}
%
\noindent
Without shrinking, QuickCheck finds large examples to the property; on average,
QuickCheck returns a value contains 70 \ttp{Int16} values!

Thus, it pays to define \ttp{shrink}.  Unfortunately, even for a simple property
and program like this one, the definition for \ttp{shrink} given in
Figure~\ref{fig:initial} produces an intractable number of potential
counterexamples to test; with the definitions and property provided (together
with QuickCheck's default \ttp{Arbitrary} instance for lists), the list of
potential counterexamples generated by \ttp{shrink} can contain more than
$10^{10}$ elements.  This is an enormous state space to search for a smaller
example and can cause the shrinking phase to take orders-of-magnitude more time
than the counterexample discovery phase.

@@

Typically, a user tries a brute-force approach to reduce the number of potential
counterexamples, for example, by truncating lists using the (\ttp{take n}) function
that returns the first $n$ elements of a list.  The trade-off is quicker
shrinking with a lower probability of finding a smaller counterexample.  For
example, redefining shrink for type \ttp{B} as follows
%

@@

%% \end{lstlisting}
%% \end{minipage}
%
\noindent
controls the exponential blowup of the state-space by taking the first ten
elements of each list, so the cross product is limited to 1000 elements.  The
downside is that potentially smaller counterexamples may be omitted.

The programmer might ask herself: ``What if we omit the need for shrinking
counterexamples altogether?''  SmallCheck is another testing framework for
Haskell that does just this.  SmallCheck is guaranteed to return the a smallest
counterexample, if one exists~\cite{sc}.  SmallCheck does this by enumerating
all possible inputs, ordered from smallest to largest, up to some user-defined
bound.  While SmallCheck is effective for testing many programs and properties
(in accordance with the \emph{small scope hypothesis}~\cite{jackson}),
counterexamples to even relatively simple properties may be infeasible to
discover due to state-space explosion.

With SmallCheck, a property is checked up to a depth set by the user, measuring
how large values are.  For example, for type \ttp{Int}, the depth $d$ is defined
as the integers in the list enumerating from $-d$ to $d$ (i.e.,
\ttp{[(-d)..d]}).  For a product type, SmallCheck must check values produced by
taking the cross-product of the type's fields at a given depth.  In
Figure~\ref{fig:initial}, we have defined instances for generating Lazy
SmallCheck (a variant of SmallCheck) tests by defining instances
for the \ttp{Serial} class.  For the property \ttp{prop} defined above (suitably
redefined to use SmallCheck's types), SmallCheck must check to depth 16 to find
the first counterexample.  Unfortunately, after a couple hours of testing, Lazy
SmallCheck is still checking values at depth six, and the number of tests scales
exponentially with respect to the depth.\footnote{All tests in this paper are
  performed on a four-core Pentium~i7 running at 2.7GHz with 8GB RAM on
  Fedora~16.  This test, and others unless noted, are performed using
  interpreted Haskell under GHC 7.4.2.}

Finally, even if we could overcome the problems described above with QuickCheck
and SmallCheck, we are left with two others.
\begin{itemize}
  \item First, when a counterexample is discovered to some property, say with
    type
%
\begin{code}
prop :: B -> Bool
\end{code}
%
\noindent
one might wish to search for different counterexamples to \ttp{prop}.  Ideally,
the testing framework would automatically add a predicate characterizing (and
perhaps generalizing) a counterexample
%
\begin{code}
cex :: B -> Bool
\end{code}
%
\noindent
as a precondition to \ttp{prop} to define a new property
\begin{code}
prop' b = not (cex b) ==> prop b
\end{code}
%
\noindent
where \ttp{==>} is implication to generate new counterexamples to the same property.

@@

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Table~\ref{table:results}, giving the results for using QuickCheck and
SmartCheck on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck,
since as we noted, it is not even feasible for an example like this).  In the
first three columns, the results of using QuickCheck to generate counterexamples
are shown.  The first column contains the results of using QuickCheck without
shrinking to provide a baseline.  The next two columns show the result of using
QuickCheck with definitions of \ttp{shrink} that limit the lists of fields of
\ttp{B} to 10 (Figure~\ref{lst:newshrink}) and 20 elements
(Figure~\ref{lst:newshrink}, with \ttp{len = 20}), respectively.  (Recall that
with the original definition for shrink from Figure~\ref{fig:initial}, some
counterexamples cause QuickCheck to not return after hours of computation.)  In
each row, we show the maximum, mean, and median number of seconds required by
QuickCheck to generate a counterexample over 100 runs.  In the last row, we show
the mean size of the final value returned, measured by the taking the cumulative
length of the lists of the fields of \ttp{B}.  Notice that while typically
counterexamples are returned in a fraction of a second, there are outliers
taking 21 seconds and 125 seconds for the two tests.  Both tests roughly cut the
size of the counterexample in half; increasing the number of potential
counterexamples can result in a significant performance penalty for this example
without much benefit.

@@

\todo{check paper against this section again, especially generalization sec.}
\paragraph{Contributions}
This paper, and the library it describes, makes the following contributions:

@@

\item an efficient counterexample reduction strategy for algebraic
  data types;

\item an approach to counterexample generalization, to present the
  user with a formula describing a set of counterexamples to a property;

\item an approach to generically strengthen a property with a precondition that
  characterizes an already-discovered counterexample for the purpose of finding
  new, dissimilar counterexamples;

\item an implementation based on generic programming that is not only more
  efficient that hand-written \ttp{shrink} instances in many cases but that
  generates the instances automatically for the user.

@@

@@

\paragraph{Outline.}
The paper precedes as follows.  We define a type class and some definitions in
Section~\ref{sec:preliminaries} used in the remainder of the paper.  In
Section~\ref{sec:shrinking}, we describe SmartCheck's reduction algorithm for
efficiently shrinking large counterexamples.  In
Section~\ref{sec:generalization}, we turn our attention to the problem of
counterexample generalization.  Here, we present two algorithms for generalizing
counterexamples and constructing formulas for describing them.  We also discuss
a method for property strengthening to discover new counterexamples.  We
describe implementation-specific details in Section~\ref{sec:implementation},
and the results of some experiments in using our tool are presented in
Section~\ref{sec:experiments}.  Related work is given in
Section~\ref{sec:related}.  Finally, concluding remarks and future work are
presented in Section~\ref{sec:conclusions}.

@@

@@

class Arbitrary a => SubTypes a where
  size     :: a -> Size
  index    :: a -> Idx -> Maybe SubVal
  replace  :: a -> Idx -> SubVal -> a
  constr   :: a -> String
  constrs  :: a -> [String]
  baseType :: a -> Bool
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value,
\item \ttp{index} returns a sub-value at an breadth-first index in a value,
\item \ttp{replace} replaces a sub-value at a particular focus,
\item \ttp{constr} returns the constructor tagging the value,
\item \ttp{constrs} returns a list of unique identifiers for the constructors
  associated with each variant of the type,
\item \ttp{baseType} returns whether the type of the value is an ``interesting
  type'', which generally means it has structure (e.g., a sum or product type).
  We discuss the purpose of base types in Section~\ref{sec:base}.
\end{itemize}

\noindent
For example, consider a binary tree type:
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with the indexes of its sub-values
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes} methods; in the following, we
show the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

baseType (3 :: Int) = True
baseType L          = False
\end{code}
%
\noindent
We will sometimes refer to the unwrapped value returned by \ttp{index a i} as
the \emph{$i$th sub-value of $a$}, so for example, \ttp{B\sub{4} L\sub{6}
  L\sub{8}} is the 4th sub-value of \ttp{tree}.  An invariant of \ttp{index} is
that for any value \ttp{a}, consider the least $i \geq 0$ such that $$\ttp{index
  a i} = \ttp{Nothing}$$ For all $0 \leq j < i$, $$\ttp{index a j} \neq
\ttp{Nothing}$$ We use this invariant as a termination case in recursive
algorithms over the sub-values of a value.

The method \ttp{replace} returns the original value for an out-of-bounds index
or an incorrectly-typed replacement value.  The \ttp{SubVal} type is an
existential data type, used as a generic container for sub-values from a
counterexample.

While \ttp{constr} returns a string representation of the constructor
associated with the specific variant passed to it, \ttp{constrs} ignores the
value and returns the same list of constructors for any value from the type.
(In our implementation, \ttp{constrs} depends on GHC Generics~\cite{generics},
which we describe in Section~\ref{sec:implementation}.  For simplicity, we omit
here Generics-specific super-class constraints on the \ttp{SubTypes} class.)

Finally, \ttp{baseType} is true for our tree type but false for a structureless
type, such as \ttp{Int}.

@@

The approach taken by \ttp{shrink} is the most naive graph search algorithm: it
simply generates a random set of potential vertexes to check in the graph.
However, it does not exploit that the original counterexample \ttp{cex} is
already in a neighborhood of other likely counterexamples.  So rather than
choose random vertexes, it may be more efficient to test other values that are
closely related to \ttp{cex}.  By restricting the search to \ttp{cex}'s
neighborhood, a smaller counterexample might be missed, and only a local minimum
is returned.  However, with a large state space, finding a global minimum by
random sampling vertexes in the graph is possible but too inefficient, as we saw
from the example in the introduction.

\subsection{Reduction Algorithm Overview}\label{sec:reduct}
The algorithm we present for efficiently searching for new counterexamples can
be thought of as a recursive breadth-first traversal over the structure of the
current counterexample value, where the value is represented as a tree.  At each
node, during the traversal, we generate arbitrary \emph{structurally} smaller
sub-values and build a new value from that, leaving the remainder of the tree
unchanged.  Intuitively, by a structurally smaller value, we mean one with fewer
constructors.  We continue until we reach a fixed-point.

@@

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes a counterexample \ttp{cex} and the property \ttp{prop}.  At
the beginning of execution, we assume that a counterexample has been obtained
somehow; our implementation relies on QuickCheck as a ``front-end'' to generate
a counterexample that the algorithm automatically shrinks.  The reduction begins
at the first sub-value of \ttp{cex}, \ttp{v}.  When the index \ttp{idx} becomes
out-of-bounds and returns \ttp{Nothing}, the algorithm terminates.  Otherwise, a
list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and strictly
smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just like
QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates successfully
larger counterexamples when generating new values with which to replace a
sub-value.

@@

\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, we assume one is generated locally).  The constant \ttp{scMaxReduce} is
the maximum number of tries to discover a new counterexample by replacing
\ttp{v} in \ttp{cex} and testing it.  The result of \ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  If no failure is found,
we move to the next sub-value of \ttp{cex} and continue.  However, if a new
smaller counterexample \ttp{cex'} is found, we start a new breadth-first
traversal of \ttp{cex'}, attempting to shrink it further.

\subsection{Reduction Algorithm Details and Optimizations}
The reduction algorithm description above omits some minor details and
optimizations we describe here.

\subsubsection{Sum-type counterexample hypothesis}
First, recall that the algorithm begins at the \emph{first} sub-value of the
counterexample rather than the 0th sub-value.  The 0th sub-value of \ttp{cex} is
\ttp{cex} itself.  No invariant of the algorithm would be violated by beginning
with the 0th sub-value; recall that it would generate strictly smaller values
than \ttp{cex}.  However, we begin with a strict sub-value because of a
hypothesis we refer to as the \emph{sum-type counterexample hypothesis}.  The
hypothesis is that in the search space of possible values, if a counterexample
of the form \ttp{C}~$\vec{x}$, where \ttp{C} is some constructor and $\vec{x}$
are its arguments, is discovered, then it is more likely to find another
``nearby'' counterexample of the form \ttp{C}~$\vec{y}$, where $\vec{y}$ are
different arguments to \ttp{C}.  For example, consider a property about
unintended variable capture over a language's parse tree represented by a sum
type with constructors for module imports, function definitions, and
global-variable assignments, respectively.  A function definition counterexample
can only be reduced to smaller function definition counterexamples, the only
construct in which variable capture is possible.

Another way to motivate the hypothesis is that searching the space of possible
counterexamples by generating \emph{any} well-typed value that is strictly
smaller than the original counterexample is something we have seen before: it is
precisely the naive approach to implementing \ttp{shrink}!

\subsubsection{Conditional properties}
A QuickCheck property may be a conditional of the form
%
\begin{code}
prop a = pre a ==> prop' a
\end{code}
%
\noindent
such that \ttp{prop} is true if the precondition \ttp{pre a} is false or
\ttp{prop' a} holds.  In practice, conditionals are used to restrict the set of
values tested; values that fail the precondition are discarded.  Thus, the call
to \ttp{pass} in the reduction algorithm only considers a potential
counterexample \ttp{cex} to have failed if it passes the property's
precondition and fails the consequent.

\subsubsection{Sub-value counterexample hypothesis}\label{sec:subval}
Sometimes, a counterexample fails a property due to a sub-value nested deep
inside the counterexample.  The rest of the value is irrelevant.  We call this
the \emph{sub-value counterexample hypothesis.}  Thus, one way to efficiently
search the space of potential counterexamples is to test a counterexample's
(well-typed) sub-values.

>>>>

\begin{abstract}
QuickCheck, developed by Klassen and Hughes, is a powerful library for automatic
test-case generation.  Because QuickCheck performs random testing, it can
discover very large counterexamples.  QuickCheck provides an interface for the
user to write \emph{shrink} functions to attempt to reduce the size of
counterexamples discovered by QuickCheck.  Implementations of \emph{shrink}
often require significant boilerplate and usually just generate smaller tests
than the discovered counterexample.  Therefore, shrinking can be inefficient and
poor at discovering new, smaller counterexamples.  In this paper, we introduce
\emph{SmartCheck}, an automatic efficient counterexample reduction algorithm.
SmartCheck reduces algebraic data using generic search heuristics to efficiently
find smaller counterexamples.  In addition to shrinking, SmartCheck generalizes
counterexamples by returning a formula that abstracts over individual
counterexamples and make it easier for the programmer to uncover multiple
counterexamples at once.  SmartCheck has been implemented for Haskell and is
freely available.
\end{abstract}

@@

\todo{define variant, tag}
\todo{note constants are user-defined arguments, other simplifications}
\todo{thread args through}
\todo{termination for each algorithm}
\todo{fromMaybe over pattern matching?}

@@

@@

While \ttp{shrink} is user-defined and can have any type-correct implementation,
it is typically defined using structural recursion for composite data types.
For example, consider the \ttp{shrink} instances given to data types \ttp{A} and
\ttp{B} in Figure~\ref{fig:initial}.\footnote{All examples and algorithms in
  this paper are presented in Haskell~\cite{haskell98}.}  For \ttp{A}, we map
the constructor \ttp{A} over a list of new \ttp{Int16}s returned by
\ttp{shrink}; for \ttp{B}, we use Haskell's list comprehension notation to
return a list of new values.  Indeed, because of the regular structure of
\ttp{shrink} implementations, the principal motivation for one of Haskell's
first generics package was for providing generic definitions for
\ttp{shrink}~\cite{syb}.

However, this straightforward approach to defining \ttp{shrink} instances may
either miss minimal counterexamples, be too inefficient to be practical, or
both.\footnote{For a recent real-world case, see the user question posed on
  StackOverflow:
  \url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}.}
For example, consider again Figure~\ref{fig:initial}.  The data type \ttp{B} is
a product type over lists of 16-bit signed integers, wrapped by a constructor.
Consider the property \ttp{prop} in the figure, stating that if the sum of each
list in a field of \ttp{B} is less than 16, then the sum of the four fields is
less than 64.  The property seems reasonable at first glance, until
one realizes that due to underflow, the property can be violated.  For example,
consider the value
%
\begin{code}
B [A (-32769)] [] [] []
\end{code}
%
\noindent
Without shrinking, QuickCheck finds large examples to the property; on average,
QuickCheck returns a value contains 70 \ttp{Int16} values!

Thus, it pays to define \ttp{shrink}.  Unfortunately, even for a simple property
and program like this one, the definition for \ttp{shrink} given in
Figure~\ref{fig:initial} produces an intractable number of potential
counterexamples to test; with the definitions and property provided (together
with QuickCheck's default \ttp{Arbitrary} instance for lists), the list of
potential counterexamples generated by \ttp{shrink} can contain more than
$10^{10}$ elements \todo{true?}.  This is an enormous state space to search for a smaller
example and can cause the shrinking phase to take orders-of-magnitude more time
than the counterexample discovery phase.

@@

Typically, a programmer tries a brute-force approach to reduce the number of
potential counterexamples; for example, she may try to truncate lists using the
(\ttp{take n}) function that returns the first $n$ elements of a list.  The
trade-off is quicker shrinking with a lower probability of finding a smaller
counterexample.  For example, redefining shrink for type \ttp{B} as follows
%

@@

%% \end{lstlisting}
%% \end{minipage}
%
\noindent
controls the exponential blowup of the input-space by taking the first ten
elements of each list, so the cross product is limited to 10,000 elements.  The
downside is that potentially smaller counterexamples may be omitted.

The programmer might ask herself: ``What if we omit the need for shrinking
counterexamples altogether?''  SmallCheck is another testing framework for
Haskell that does just this.  SmallCheck is guaranteed to return a smallest
counterexample, if one exists~\cite{sc}.  SmallCheck does this by enumerating
all possible inputs, ordered from smallest to largest, up to some user-defined
bound.  While SmallCheck is effective for testing many programs and properties
(in accordance with the \emph{small scope hypothesis}~\cite{jackson}),
counterexamples to even relatively simple properties may be infeasible to
discover due to input-space explosion.

\begin{figure}
  \begin{code}
instance Serial Int16 where
  series d = drawnFrom [(-d')..d']
    where d' = fromIntegral d

instance Serial A where series = cons1 A
instance Serial B where series = cons4 B
  \end{code}
  \caption{SmallCheck instances for the data types in
    Figure~\ref{fig:initial}.}
  \label{fig:smallshrink}
\end{figure}

With SmallCheck, a property is checked up to a depth set by the user that
measures how large values are.  For example, for type \ttp{Int}, the depth $d$
is defined as the integers in the list enumerating from $-d$ to $d$ (i.e.,
\ttp{[(-d)..d]}).  For a product type, SmallCheck must check values produced by
taking the cross-product of the type's fields at a given depth.
Figure~\ref{fig:smallshrink} contains functions for generating Lazy SmallCheck
(a variant of SmallCheck) tests by defining instances for the \ttp{Serial}
class.  For the property \ttp{prop} defined above (suitably redefined to use
SmallCheck's types), SmallCheck must check to depth 16 to find the first
counterexample.  Unfortunately, after a couple hours of testing, Lazy SmallCheck
is still checking values at depth six, and the number of tests scales
exponentially with respect to the depth.\footnote{All tests in this paper are
  performed on a four-core Pentium~i7 running at 2.7GHz with 8GB RAM on
  Fedora~16.  This test, and others unless noted, are performed using
  interpreted Haskell under GHC 7.4.2.\todo{redo tests?}}

Finally, even if we could overcome the problems described above with QuickCheck
and SmallCheck, we are left with two others.
\begin{itemize}
  \item First, after discovering a counterexample to some property, say with
    type
%
\begin{code}
prop :: B -> Bool
\end{code}
%
\noindent
one might wish to search for different counterexamples to \ttp{prop}.  Ideally,
the testing framework would automatically add a predicate characterizing (and
perhaps generalizing) a counterexample
%
\begin{code}
cex :: B -> Bool
\end{code}
%
\noindent
as a precondition to \ttp{prop} to define a new property
\begin{code}
prop' b = not (cex b) ==> prop b
\end{code}
%
\noindent
where \ttp{==>} is QuickCheck's implication operator that discards tests that
violate the precondition.

@@

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Table~\ref{table:results}, giving the results for using QuickCheck and
SmartCheck on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck,
since as we noted, it is not even feasible for an example like this).  In the
first three columns, the results of using QuickCheck to generate counterexamples
are shown.  The first column contains the results of using QuickCheck without
shrinking to provide a baseline.  The next two columns show the result of using
QuickCheck with definitions of \ttp{shrink} that limit the lists of fields of
\ttp{B} to 10 (Figure~\ref{lst:newshrink}) and 20 elements
(Figure~\ref{lst:newshrink}, with \ttp{len = 20}), respectively.  (With the
original definition for shrink from Figure~\ref{fig:initial}, some
counterexamples cause QuickCheck to not return during the shrinking phase after
hours of computation, so we omit these.)

  In each row, we show the maximum, mean, and median
number of seconds required by QuickCheck to generate a counterexample over 100
runs.  In the last row, we show the mean size of the final value returned,
measured by the taking the cumulative length of the lists of the fields of
\ttp{B}.  Notice that while typically counterexamples are returned in a fraction
of a second, there are outliers taking 21 seconds and 125 seconds for the two
tests.  Both tests roughly cut the size of the counterexample in half;
increasing the number of potential counterexamples can result in a significant
performance penalty for this example without much benefit.

@@

\todo{check paper against this section again, especially generalization sec.}

\paragraph{Contributions}
This paper, and the library it describes, makes the following contributions:

@@

\item an efficient counterexample reduction strategy for algebraic
  data types (Section~\ref{sec:shrinking});

\item an approach to counterexample generalization, to present the user with a
  formula describing a set of counterexamples to a property
  (Section~\ref{sec:generalization});

\item an approach to generically strengthen a property with a precondition that
  characterizes an already-discovered counterexample for the purpose of finding
  new, dissimilar counterexamples(Section~\ref{sec:generalization});

\item an implementation based on generic programming that is not only more
  efficient that hand-written \ttp{shrink} instances in many cases but that
  generates the instances automatically for the user
  (Section~\ref{sec:implementation}).

@@

\noindent
In Section~\ref{experiments}, we discuss some of our experiences with using
SmartCheck, including checking properties from Xmonad.

@@

%% \paragraph{Outline.}
%% The paper precedes as follows.  We define a type class and some definitions in
%% Section~\ref{sec:preliminaries} used in the remainder of the paper.  In
%% Section~\ref{sec:shrinking}, we describe SmartCheck's reduction algorithm for
%% efficiently shrinking large counterexamples.  In
%% Section~\ref{sec:generalization}, we turn our attention to the problem of
%% counterexample generalization.  Here, we present two algorithms for generalizing
%% counterexamples and constructing formulas for describing them.  We also discuss
%% a method for property strengthening to discover new counterexamples.  We
%% describe implementation-specific details in Section~\ref{sec:implementation},
%% and the results of some experiments in using our tool are presented in
%% Section~\ref{sec:experiments}.  Related work is given in
%% Section~\ref{sec:related}.  Finally, concluding remarks and future work are
%% presented in Section~\ref{sec:conclusions}.

@@

We focus on \emph{algebraic data types}~\cite{}.  Algebraic data types are
represented as ``sums of products''.  We sometimes refer to the elements of a
sum type as a \emph{variant} that is \emph{tagged} by a constructor.  For
example, the type
%
\begin{code}
Maybe a = Just a | Nothing
\end{code}
%
\noindent
contains two variants, tagged by the constructors \ttp{Just} and \ttp{Nothing},
respectively.

@@

class Arbitrary a => SubTypes a where
  size     :: a -> Size
  index    :: a -> Idx -> Maybe SubVal
  replace  :: a -> Idx -> SubVal -> a
  constr   :: a -> String
  constrs  :: a -> [String]
  baseType :: a -> Bool
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value,
\item \ttp{index} returns a \emph{sub-value} at an breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{baseType} returns whether the type of the value is an ``interesting
  type'', which generally means it has structure (e.g., a sum or product type).
  We discuss the purpose of base types in Section~\ref{sec:base}.
\end{itemize}

\noindent
For example, consider a binary tree type:
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with indexes in a breadth-first order:
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes} methods; in the following, we
show the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

baseType (3 :: Int) = True
baseType L          = False
\end{code}
%
\noindent
The \ttp{SubVal} type is an existential data type, used as a generic container
for sub-values from a counterexample.  We will sometimes refer to the unwrapped
value returned by \ttp{index a i} as the \emph{$i$th sub-value of $a$}, so for
example, (\ttp{B\sub{4} L\sub{6} L\sub{8}}) is the 4th sub-value of \ttp{tree}.
An invariant of \ttp{index} is that for any value \ttp{a}, and for the smallest
$i \geq 0$ such that
%
\begin{code}
index a i == Nothing
\end{code}
%
\noindent
then for all $0 \leq j < i$,
%
\begin{code}
index a j /= Nothing
\end{code}
\noindent
We use this invariant as a termination case in recursive algorithms over the
sub-values of a value.

In our implementation, \ttp{constr} and \ttp{constrs} depends on GHC
Generics~\cite{generics}, which we describe in Section~\ref{sec:implementation}.
For simplicity, we omit here Generics-specific super-class constraints on the
\ttp{SubTypes} class.

%% Finally, \ttp{baseType} is true for our tree type but false for a
%% structureless type, such as \ttp{Int}.

@@

The approach taken by \ttp{shrink} is the most naive graph search algorithm: it
simply generates a random set of potential vertexes to check in the graph.
However, it does not exploit that the original counterexample \ttp{cex} is
already in a neighborhood of other likely counterexamples.  So rather than
choose random vertexes, it may be more efficient to test other values that are
closely related to \ttp{cex}.  By restricting the search to \ttp{cex}'s
neighborhood, a smaller counterexample might be missed, and only a local minimum
is returned.  However, with a large set of values, finding a global minimum by
random sampling vertexes in the graph is possible but too inefficient, as we saw
from the example in the introduction.

\subsection{Reduction Algorithm Overview}\label{sec:reduct}
The algorithm we present for efficiently searching for new counterexamples can
be thought of as a recursive breadth-first traversal over the structure of the
current counterexample value, where the value is represented as a tree.  At each
node, during the traversal, we generate arbitrary \emph{structurally} smaller
sub-values and build a new value from that, leaving the remainder of the tree
unchanged.  Intuitively, by a \emph{structurally smaller value}, we mean one
with fewer constructors.  We continue until we reach a fixed-point.

@@

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes a counterexample \ttp{cex} and the property \ttp{prop}.  At
the beginning of execution, we assume that a counterexample has been obtained
somehow; our implementation relies on QuickCheck as a ``front-end'' to generate
a counterexample that the algorithm automatically shrinks.  The reduction begins
at the first proper sub-value of \ttp{cex}; call it \ttp{v}.  When the index
\ttp{idx} becomes out-of-bounds and returns \ttp{Nothing}, the algorithm
terminates.  Otherwise, a list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successfully larger counterexamples when generating new values with which to
replace a sub-value.

@@

\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, let us assume one is generated locally).  The constant
\ttp{scMaxReduce} is the maximum number of tries to discover a new
counterexample by replacing \ttp{v} in \ttp{cex} and testing it.  The result of
\ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  The property may be a
conditional, in which case the value must pass the precondition as well as the
consequent for \ttp{pass} to return \ttp{True}.  If no failure is found, we move
to the next sub-value of \ttp{cex} and continue.  However, if a new smaller
counterexample \ttp{cex'} is found, we start a new breadth-first traversal of
\ttp{cex'}, attempting to shrink it further.  The algorithm is guaranteed to
terminate: intuitively, the measure for the function is that either the index
increases or the size of the counterexample being evaluated decreases.

\subsection{Reduction Algorithm Details}
Having describe the reduction algorithm, there are two important details about
its design we describe below.

\subsubsection{Variant Counterexample Hypothesis}
A motivation for the design of the reduction algorithm is something we cal the
\emph{variant counterexample hypothesis}.  Recall that the algorithm begins at
the \emph{first} sub-value of the counterexample rather than the 0th sub-value.
The 0th sub-value of \ttp{cex} is \ttp{cex} itself.  No invariant of the
algorithm would be violated by beginning with the 0th sub-value, and in
particular, the algorithm would still terminate.  However, we begin with a
strict sub-value because of the hypothesis.  The hypothesis is that in the
search space of possible values from a given type \ttp{T}, starting with a
particular counterexample \ttp{cex}, we are more likely to find other
counterexamples from the same variant as \ttp{cex}.  For example, consider a
property about unintended variable capture over a language's parse tree
represented by a sum type with constructors for module imports, function
definitions, and global-variable assignments, respectively.  A function
definition counterexample can only be reduced to smaller function definition
counterexamples, the only construct in which variable capture is possible.

%% Another way to motivate the hypothesis is that searching the space of
%% possible counterexamples by generating \emph{any} well-typed value that is
%% strictly smaller than the original counterexample is something we have seen
%% before: it is precisely the naive approach to implementing \ttp{shrink}!

%% \subsubsection{Conditional Properties}
%% A QuickCheck property may be a conditional of the form
%% %
%% \begin{code}
%% prop a = pre a ==> prop' a
%% \end{code}
%% %
%% \noindent
%% such that \ttp{prop} is true if the precondition (\ttp{pre a}) is false or
%% (\ttp{prop' a}) holds.  In practice, conditionals are used to restrict the set
%% of values tested; values that fail the precondition are discarded.  Thus, the
%% call to \ttp{pass} in the reduction algorithm only considers a potential
%% counterexample \ttp{cex} to have failed if it passes the property's precondition
%% and fails the consequent.

\subsubsection{Base Types}\label{sec:base}

The number of constructors in an algebraic data value primarily determines how
large and understandable it is.  In our experience, the size of fields that come
from primitive types, like \ttp{Int} or \ttp{Char}, have less of a role in
understanding a counterexample.  However, a testing algorithm might spend
significant time testing new values in which primitive types are reduces.

Consequently, we distinguish a class of types referred to as \emph{base types}.
If the reduction algorithm encounters a base type, it is ignored.  By default,
this includes numeric types (e.g., \ttp{Int}, \ttp{Word}, \ttp{Float}, etc.,
\ttp{Bool}, and \ttp{Char}.  We also treat recursive container types like lists
specially: lists of base types are considered to be base types themselves.  This
prevents the algorithm from analyzing, for example, strings.  A list of a
non-base-types is shrunk by the algorithm.
